Detecting distributional differences between temporal granularities for exploratory time series analysis

### 2021-08-18 paper/main.Rmd
workingon intro

Exploratory data analysis, as coined by John W. Tukey [@Tukey1977-jx] involves many iterations of finding structures and patterns that allow the data to be informative. With temporal data available at finer scales, exploring time series data can become overwhelming with so many possible cyclic temporal granularities [@Bettini1998-ed, @Gupta2021-hd], which are temporal deconstructions that represent cyclic repetitions in time, e.g. hour-of-day, day-of-month, or any public holidays. These granularities form ordered (for eg. day-of-week where Tue is always followed by Wed, which again is followed by Thu and so on) or unordered categorical variables. Therefore, exploring univariate time series data amounts to exploring the distribution of the measured variable across different categories of the cyclic granularities. \noindent Take the example of the electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. Figure \ref{fig:onegran-new} shows the distribution of energy usage of one household (id 2) across cyclic granularity a) hour-of-day and b) month-of-year. Figure \ref{fig:onegran-new}(a) shows that energy consumption is higher during the morning hours (5-8) when members in the household wake up and then again in the evening hours (17-20) possibly when members get back from work with maximum variation (large interquartile range) in behavior in the afternoon hours (12-16). Figure \ref{fig:onegran-new}(b) shows the distribution of energy consumption across months January to June. The median and quartile deviation of energy usage in Jan and Feb are generally on a much higher side, possibly due to the usage of air conditioners (Jan, Feb are peak summer in Australia), however, for other months (Mar-Jun, autumn and winter), the smaller median and quartile deviation indicate a more consistent behavior. It might also imply that this household does not use as much heater as compared to air conditioner. A lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. Potentially many such displays could be drawn across day-of-week, day-of-month, weekday/weekend, or any other chosen cyclic granularities of interest. However, all of them would not be interesting to discern important patterns in energy usage. Only those displays, which have "significant" distributional differences between categories of the cyclic granularity would be informative.

>>>>

<!--Exploratory data analysis, as coined by John W. Tukey [@Tukey1977-jx] involves many iterations of finding structures and patterns that allow the data to be informative.-->

Cyclic temporal granularities [@Bettini1998-ed, @Gupta2021-hd] are theoretical temporal deconstructions that define cyclic repetitions in time, e.g. hour-of-day, day-of-month, or regularly scheduled public holidays. These granularities form ordered or unordered categorical variables. An example of an ordered granularity is day-of-week, where Tuesday is always followed by Wednesday, and so on. An unordered granularity might be XXX. Thus we can consider that exploring univariate time series using granularities to be examining the distribution of the measured variable across different categories of the cyclic granularities.

As an example, Figure \ref{fig:onegran-new} shows electricity smart meter data for four households in Melbourne, Australia [@wang2020calendar]. The distribution of energy usage of one household (id 2) across cyclic granularity a) hour-of-day and b) month-of-year. Figure \ref{fig:onegran-new}(a) shows that energy consumption is higher during the morning hours (5-8) when members in the household wake up and then again in the evening hours (17-20) possibly when members get back from work with maximum variation (large interquartile range) in behavior in the afternoon hours (12-16). Figure \ref{fig:onegran-new}(b) shows the distribution of energy consumption across months January to June. The median and quartile deviation of energy usage in Jan and Feb are generally on a much higher side, possibly due to the usage of air conditioners (Jan, Feb are peak summer in Australia), however, for other months (Mar-Jun, autumn and winter), the smaller median and quartile deviation indicate a more consistent behavior. It might also imply that this household does not use as much heater as compared to air conditioner. A lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. Potentially many such displays could be drawn across day-of-week, day-of-month, weekday/weekend, or any other chosen cyclic granularities of interest. However, all of them would not be interesting to discern important patterns in energy usage. Only those displays, which have "significant" distributional differences between categories of the cyclic granularity would be informative.
